---
title: Loess Regression
layout: default
sitemap:
    priority: 1
    changefreq: 'monthly'
    lastmod: 2015-10-29T12:49:30-05:00
---
> Loess Regression is the most common method used to smoothen a volatile time series. It is a non-parametric methods where least squares regression is performed in localized subsets, which makes it a suitable candidate for smoothing any numerical vector.

## Introduction
_Loess_ short for Local Regression is a non-parametric approach that fits multiple regressions in local neighborhood. This can be particularly resourceful, if you know that your $X$ variables are bound within a range. 

Loess regression can be applied using the `loess()` on a numerical vector to smoothen it and to predict the Y locally (i.e, within the trained values of $Xs$). The size of the neighborhood can be controlled using the `span` argument, which ranges between 0 to 1. It controls the degree of smoothing. So, the greater the value of `span`, more smooth is the fitted curve. 

The predictor variable can just be indices from 1 to number of observations in the absence of explanatory variables. If other explanatory variables are available, they can be used as well (maximum of 4).

## Example
For this example we will try to locally regress and smooth the _median duration of unemployment_ based on the `economics` dataset from `ggplot2` package. We consider only the first 80 rows for this analysis, so it is easier to observe the degree of smoothing in the graphs below.

```{r, eval=FALSE}
data(economics, package="ggplot2")  # load data
economics$index <- 1:nrow(economics)  # create index variable
economics <- economics[1:80, ]  # retail 80rows for better graphical understanding
loessMod10 <- loess(uempmed ~ index, data=economics, span=0.10) # 10% smoothing span
loessMod25 <- loess(uempmed ~ index, data=economics, span=0.25) # 25% smoothing span
loessMod50 <- loess(uempmed ~ index, data=economics, span=0.50) # 50% smoothing span
```

## Predict Loess
```{r, eval=FALSE}
# get smoothed output
smoothed10 <- predict(loessMod10) 
smoothed25 <- predict(loessMod25) 
smoothed50 <- predict(loessMod50) 
```

```{r, echo = FALSE}
embed_png("screenshots/Loess-Smoothing.png", dpi = 220)
```

From above plot, you would notice that as the span increases, the smoothing of the curve also increases.

##### Code for Plot
```{r, eval=FALSE}
# Plot it
plot(economics$uempmed, x=economics$date, type="l", main="Loess Smoothing and Prediction", xlab="Date", ylab="Unemployment (Median)")
lines(smoothed10, x=economics$date, col="red")
lines(smoothed25, x=economics$date, col="green")
lines(smoothed50, x=economics$date, col="blue")
```

## Finding the optimal smoothing span

As the smoothing `span` changes, the accuracy of the fitted curve also changes. If your intent is to minimize the error, the `optim()` can be used to find that value of `span`, that minimizes the _Sum of Squared Errors (SSE)_. For this case, it is graphically intuitive that lower SSE will likely be achieved at lower values of `span`, but for more challenging cases, `optimizing` span could help.

To implement `optim()`, we define the function that computes the _SSE_. An error handling mechanism is needed to address very low values of `span` and cases where the non-numerics are produced. The simulated annealing method (`SANN`) is implemented here to find the `span` that gives minimal _SSE_.
The `par` argument specifies the first value of the `span` at which `optim()` will begin the search.

```{r, eval=FALSE}
# define function that returns the SSE
calcSSE <- function(x){
  loessMod <- try(loess(uempmed ~ index, data=economics, span=x), silent=T)
  res <- try(loessMod$residuals, silent=T)
  if(class(res)!="try-error"){
    if((sum(res, na.rm=T) > 0)){
      sse <- sum(res^2)  
    }
  }else{
    sse <- 99999
  }
  return(sse)
}

# Run optim to find span that gives min SSE, starting at 0.5
optim(par=c(0.5), calcSSE, method="SANN")
#> $par
#> [1] 0.05433545
#> 
#> $value
#> [1] 3.85753e-28
#> 
#> $counts
#> function gradient 
#>    10000       NA 
#> 
#> $convergence
#> [1] 0
#> 
#> $message
#> NULL
```

For this case, the best value of `span` turns out to be `0.05433` and the minimum SSE achieved is `3.85e-28`.


