---
title: Poisson and Negative Binomial Regression
layout: default
sitemap:
    priority: 1
    changefreq: 'monthly'
    lastmod: 2015-10-29T12:49:30-05:00
---
# Poisson and Negative Binomial Regression

> _Poisson regression_ models count variables that assumes poisson distribution. When the count variable is over dispersed, having to much variation, _Negative Binomial_ regression is more suitable.

## Introduction

A count variable is something that can take only non-negative integer values. Some examples of count variables could be:
1. Number of vehicles manufactured.
1. Number of phone calls arriving at a call center.
1. Number of patents granted.

## How to Implement Poisson Regression? 

Poisson regression can be implemented in a similar manner as other _Generalised Linear Models (GLMs)_, by adjusting the family argument to `poisson`.

```{r, eval=FALSE}
library (MASS)
poissonModel <- glm(countResponse ~ pred1 + pred2, family="poisson", data=inputData) # poisson Model
summary (poissonModel) # model summary
predict(poissonModel, newdata, type="response") # predict on new data 
```

## How to Implement Negative Binomial Regression?

```{r, eval=FALSE}
library (MASS)
negBinomModel <- glm.nb(countResponse ~ pred1 + pred2, data = inputData)) # negative Binomial model
summary (negBinomModel) # Model summary
predict (negBinomModel, newdata, type="response") # predict on new data 
```
