---
title: Robust Regression
layout: default
sitemap:
    priority: 1
    changefreq: 'monthly'
    lastmod: 2015-10-29T12:49:30-05:00
---
# Robust Regression

> Robust regression can be used in any situation where OLS regression can be applied. It generally gives better accuracies over OLS because it uses a weighting mechanism to weigh down the influential observations. It is particularly resourceful when there are no compelling reasons to exclude outliers in your data.

Robust regression can be implemented using the `rlm()` function in MASS package. The outliers can be weighted down differently based on `psi.huber`, `psi.hampel` and `psi.bisquare` methods specified by the `psi` argument.

## How To Specify A Robust Regression Model

```{r, eval=FALSE}
library(MASS)
rlm_mod <- rlm(stack.loss ~ ., stackloss, psi = psi.bisquare)  # robust reg model
summary(rlm_mod)
#> Call: rlm(formula = stack.loss ~ ., data = stackloss)
#> Residuals:
#>      Min       1Q   Median       3Q      Max 
#> -8.91753 -1.73127  0.06187  1.54306  6.50163 
#> 
#> Coefficients:
#>             Value    Std. Error t value 
#> (Intercept) -41.0265   9.8073    -4.1832
#> Air.Flow      0.8294   0.1112     7.4597
#> Water.Temp    0.9261   0.3034     3.0524
#> Acid.Conc.   -0.1278   0.1289    -0.9922
#> 
#> Residual standard error: 2.441 on 17 degrees of freedom
```

## Compare Performance of `rlm()` with `lm()`

Lets build the equivalent `lm()` model so we can compare the errors against the respective fitted values.

```{r, eval=FALSE}
lm_mod <- lm(stack.loss ~ ., stackloss)  # lm reg model
```

#### Calculate the Errors
```{r, eval=FALSE}
# Errors from lm() model
DMwR::regr.eval(stackloss$stack.loss, lm_mod$fitted.values)
#>       mae       mse      rmse      mape 
#> 2.3666202 8.5157125 2.9181694 0.1458878

# Errors from rlm() model
DMwR::regr.eval(stackloss$stack.loss, rlm_mod$fitted.values)
#>       mae       mse      rmse      mape 
#> 2.1952232 9.0735283 3.0122298 0.1317191
```

As expected, the errors from the robust regression model is lesser than the linear regression model.